Emittance preservation of an electron bunch in a loaded quasi-linear plasma wakefield

### 2017-08-24 Tex/Paper.tex
Fixed typo
Text cleanup

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and has been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
will excite a strong longitudinal e-field that can be loaded by a trailing witness beam, which can achieve a high
acceleration gradient. Acceleration of an electron beam by an electron drive bunch has been demonstrated experimentally
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. The AWAKE experiment at CERN is a proof of concept
proton driven plasma wakefield accelerator \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam with a minimal increase
in energy spread and emittance. In the well described linear case where the beam density $n_{b}$ is much smaller than
the plasma density $n_{0}$ the plasma causes emittance growth in the beam. A finite length beam will also see a varying
transverse wakefield causing increasing energy spread \cite{katsouleas:1987}. In the non-linear regime, where
$n_{b} > n_{0}$, the drive beam sees an unvarying transverse field due to a bubble forming in the plasma behind the
drive beam. The bubble is formed by the transverse oscillations of the plasma electrons which move near uniformly as a
function of distance from the drive beam forming a sheet around an evacuated area. Since the much heavier plasma ions
have no significant movement within the relevant time frame, these remaining ion channel creates a focusing force that
scales linearly with the radius within this bubble producing a angularly symmetric focusing force
\cite{lu:2006-1, lu:2006}. This effect also avoid the issue of increasing energy spread due to transverse beam loading.
Increase in energy spread is still dependent on beam loading of the longitudinal e-field, $E_{z}$. This effect can be
negated by specially shaped beams \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not
a topic explored in this paper.

@@

This de-phasing problem can be minimised by instead using a proton drive bunch \cite{caldwell:2009}. Proton beams, like
for instance the LHC and SPS beams at CERN, are ideal candidates for such an accelerator; however, they are orders of
magnitude longer than the plasma wavelengths needed for such applications. This issue is resolved by letting the proton
beam undergo self-modulation before injecting the witness beam into one of the buckets in the modulated structure. This
self-modulation instability is caused by the transverse fields generated by the beam acting upon the beam itself,
causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency is close to that of the
plasma, leaving a train of proton bunches along the beam axis.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam as its driver. The first run
of the experiment is set up with a $10\unit{m}$ rubidium plasma cell, with a nominal plasma density of
$7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. The corresponding plasma wavelength is $1.26\unit{mm}$. The plasma
density is matched to the SPS proton beam such that $k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the
plasma wave number. The aim of the first run is to demonstrate self-modulation of the proton beam, and in 2018 to
sample the wake field with a long electron bunch. The study presented here is for run two, which aims to demonstrate
acceleration of a short electron bunch producing at high energy and low emittance and low energy spread.

@@

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage. The electron beam will be injected into the modulated
proton beam before stage two, where acceleration will occur. As the $e_z$ field will decrease due to the gap between the
two cells, it is desirable to keep this as short as possible \cite{adli:2016}.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors that
may affect this, we have tried several approaches to create a stable drive beam structure based on previous SMI studies.

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to the
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe}^{-1} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$,
and the electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter
scans with this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low
energy spread \cite{adli:2016, berglyd_olsen:2016}.

In order to evaluate the quality of the beam, we also needed to study emittance evolution. However, full PIC codes like
Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}.
This is a know issue with the Yee EMF solver, which causes the phase velocity of electromagnetic fields to be lower than
$c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing the Yee solver with a
solver designed by Lehe \cite{lehe:2013}. These simulations showed that this numerical effect is still prominent in the
high density regions of the electron beam, making it difficult to distinguish the physics from the numerical error.

Quasi-static PIC codes do not suffer this problem, so in order to study the emittance evolution of the beam we instead
turned to the recently released open source version QuickPIC developed by UCLA. QuickPIC a fully relativistic 3D PIC
code \cite{huang:2006, an:2013}.

@@

In these simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in AWAKE. The baseline AWAKE drive beam contains $3\nexp{11}$ protons
\cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded in the
middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during self-modulation. The
needed gap between the two plasma cells also contributes to a decreased density of protons near the axis, and some of
the field is drained by other protons seeing an accelerating phase within the drive beam. In total, the expected peak
$E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in the second cell
\cite{awake_collaboration:2016}. A single proton beam of $1.46\nexp{10}$, or $2.34\unit{nC}$ and $7\unit{kA}$, is needed
to drive an $E_{z}$ field of a comparable magnitude.

The single bunch approach was chosen to reduce simulation time for the 3D simulations, and to create a stable
environment for the witness beam. We are not considering the evolution of the proton beam in this test setup, so to
prevent the proton beam from evolving radially, we also increased its mass.

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam.
This condition is replicated in out single bunch case as the peak density of the drive beam is $0.83\cdot n_{0}$. The
single bunch setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size
$\sigma_{r} = 200\unit{\mu m}$. The length of the bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. AWAKE baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$
\cite{gschwendtner:2016}. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. AWAKE Run 2 will likely require a higher initial energy, but a $W_{eb} = 30-50\unit{MeV}$ is
sufficient to minimise de-phasing as the drift $\Delta\xi \propto \gamma^{-2}$ \cite{berglyd_olsen:2015}. For this case
we have eliminated it entirely in order to reduce the number of variables in our parameter scans.

Earlier simulations with witness beam $\sigma_{r} = 100\unit{\mu m}$ caused the beam radius to undergo rapid transverse
evolution as the beam enters the plasma, reducing its $\sigma_{r}$ from $100\unit{\mu m}$ to $< 10\unit{\mu m}$
\cite{berglyd_olsen:2016}. This is due to a mismatch between beam emittance and the plasma density. This relation is
given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. In these simulations we have used matched beams for emittances between
$2$ and $6\unit{\mu m}$, corresponding to $\sigma_{r}$ ranging from $5.25\unit{\mu m}$ to $9.09\unit{\mu m}$.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent fields throughout the plasma stage. The field generated by the proton drive bunch is seen as the dotted
line in figure \ref{Fig:BeamLoading}. With a proton beam density $n_{pb} \simeq n_{0}$, we are in the quasi-linear
regime but near non-linear \cite{rosenzweig:2010}. There is some depletion of plasma electrons, but not enough to form a
bubble. The dashed green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis plasma density
has a depletion of $67\%$.

@@

The witness bunch generates its own wakefield, which in the accelerating phase cancels out the $E_{z}$ field generated
by the drive bunch. For a finite length bunch this causes the particles in the tail of the witness bunch to be
accelerated less than those at the front resulting in high energy spread in exchange for high energy transfer
\cite{van_der_meer:1985}. Since $E_{z} \propto \rho_{b}$, a beam profile $\rho_{b}(\xi)$ that perfectly flattens the
accelerating field within the witness beam, reduces energy spread to zero. The ideal shape is triangular or trapezoidal
with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    %~ \includegraphics[height=22.2mm,trim={ 0mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation01}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation02}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  0mm 0mm},clip]{figures/beamFilamentation03}
    %~ \includegraphics[height=22.2mm,trim={10mm 9.5mm 15mm 0mm},clip]{figures/beamOffsetMean}
    %~ \\
    %~ \includegraphics[height=26.0mm,trim={ 0mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation04}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation05}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  0mm 2mm},clip]{figures/beamFilamentation06}
    %~ \includegraphics[height=26.0mm,trim={10mm 0.1mm 15mm 2mm},clip]{figures/beamOffsetRMS}
    \caption{\label{Fig:BeamFilament} The left plots show the transverse phase space of the electron beam at different
        plasma position. The Right side shows the macro particle mean position (top) and RMS spread (bottom). The blue
        particles and lines represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. The red
        particles and lines represent particles with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

For Gaussian beams a reasonable flat field and consequently low energy spread can be achieved by controlling the charge
density. In our case, with a matched beam to the plasma density defined by equation \ref{EQ:Matched}, we are limited on
how much total charge we can accelerate. The test case shown in figures \ref{Fig:BeamLoading} and
\ref{Fig:PlasmaDenTWake} has a total charge of $100\unit{pC}$ with a $\sigma_{r}=5.23\unit{\mu m}$ matching an initial
normalised emittance $\epsilon_{0} = 2\unit{\mu m}$.

As illustrated in figure \ref{Fig:BeamPS}, our base case of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam, as
expected, shows a difference in energy transfer to the tail of the beam compared to the centre where the $E_{z}$-field
is nearly flat. Also the front of the beam has a large tail in momentum space corresponding to a smaller acceleration
gradient. The positioning of the witness beam is chosen to put the bulk of the charge as close to the peak accelerating
field as possible, but prioritising the tail end over the front as the front is in any case in the quasi-linear regime.

Due to the high charge density resulting from a very narrow beam, the witness beam's own wakefield reaches the full
non-linear regime with $n_{eb} \approx 35\times n_{0}$. The wakefield is driven by the beam's own front. The space
charge rapidly becomes high enough to start expel plasma electrons from the beam axis, and form the characteristic
electrons sheet that defines the blowout regime. As a result the plasma region reaches full depletion close to the peak
of the electron beam, which in turn generates strong focusing fields, with gradients of $20\unit{kT/m}$ near the beam
axis (see figure \ref{Fig:PlasmaDenTWake}).

@@

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved.

@@

Our configuration also shows some robustness to electron beam transverse offset as the bubble generated by the witness
beam wakefield follows the beam. The head of the beam does not benefit from this effect, and we see some defocusing in
this region (see bottom, right plot of figure \ref{Fig:BeamFilament}). However, since the proton wakefield creates a
quasi linear condition, we still see the head of the beam stabilising (see top, right plot of figure
\ref{Fig:BeamFilament}). The off-axis case has a larger initial emittance growth (left side plot of figure
\ref{Fig:BeamFilament}), but the growth levels off after the first few metres and stays constant for the remainder of
the plasma accelerator section (figure \ref{Fig:BeamEmitt}).

@@

For accelerator applications in general, and also for Run II of the AWAKE experiment, it is desirable to maximise the
charge that can be successfully accelerated in the witness bunch. However, since longitudinally constant accelerating
fields depend on both the position and the charge density of the witness beam \cite{katsouleas:1987, tzoufras:2009}.

@@

A higher charge beam will generate a non-linear wakefield sooner, which in turns ensures that a smaller portion of the
beam experiences the quasi-linear region with emittance growth. As figure \ref{Fig:BeamQAbs} illustrates, a higher
charge also results in a lower energy gain.

>>>>

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and has been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
will excite a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading, the
witness beam can have a high acceleration gradient. Acceleration of an electron beam by an electron drive bunch has been
demonstrated experimentally \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. The AWAKE experiment at
CERN is a proof of concept proton driven plasma wakefield accelerator \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam with a minimal increase
in energy spread and emittance. In the well described linear case where the beam density $n_{b}$ is much smaller than
the plasma density $n_{0}$ the plasma causes emittance growth in the beam. A finite length beam will also see a varying
transverse wakefield causing increasing energy spread \cite{katsouleas:1987}. In the non-linear regime, where
$n_{b} > n_{0}$, the drive beam sees an unvarying transverse field due to a bubble forming in the plasma behind the
drive beam. The bubble is formed by the transverse oscillations of the plasma electrons which move near uniformly as a
function of distance from the drive beam, forming a sheet around an evacuated area. The much heavier plasma ions have no
significant movement within the relevant time frame, and form an ion channel creating a focusing force that scales
linearly with the radius. This produces an angularly symmetric focusing force \cite{lu:2006-1, lu:2006}. This effect
also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread is still
dependent on beam loading of the longitudinal e-field, $E_{z}$. Non-uniform beam loading can be negated by specially
shaped beams that optimally flattens the field \cite{katsouleas:1987, tzoufras:2009}, but this is difficult
experimentally and is not a topic explored in this paper.

@@

This de-phasing problem can be minimised by instead using a proton drive bunch \cite{caldwell:2009}. Proton beams, like
for instance the LHC and SPS beams at CERN, are ideal candidates for such an accelerator. However, they are orders of
magnitude longer than the plasma wavelengths needed for such applications. This issue is resolved by letting the proton
beam undergo self-modulation before injecting the witness beam into one of the buckets in the modulated structure. This
self-modulation instability is caused by the transverse fields generated by the beam acting upon the beam itself,
causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency is close to that of the
plasma, and leaves a train of proton bunches along the beam axis with a surrounding halo.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam as its driver. The first run
of the experiment is set up with a $10\unit{m}$ rubidium vapour cell. The rubidium vapour is ionised by a
$780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$. The nominal plasma density of
the AWAKE experiment is $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density corresponds to a plasma
wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam such that
$k_{pe}\sigma_{r} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first phase of Run 1
of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the wake field with a long
electron bunch. The study presented here is for Run 2, which aims to demonstrate acceleration of a short electron bunch
to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will be injected into the modulated proton beam before stage two,
where acceleration will occur. As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
keep this as short as possible \cite{adli:2016}.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors that
may affect this, we have tried several approaches to create a stable drive beam structure based on previous self-
modulation studies.

Our first approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations. The proton beam was pre-modulated by a clipped cosine function to its
longitudinal density profile, with a period matching $\lambda_{pe}$. Its transverse profile was kept as a Gaussian with
$\sigma_{r} = k_{pe}^{-1} = 200\unit{\mu m}$. The length of the simulation beam was limited to $26\cdot\lambda_{pe}$,
and the electron beam injected after the 20th micro-bunch \cite{berglyd_olsen:2015}. We performed several parameter
scans with this setup, testing for the the maximum beam charge that could be accelerated with a high energy gain and low
energy spread \cite{adli:2016, berglyd_olsen:2016}.

In order to evaluate the quality of the beam, we also needed to study emittance evolution. However, full PIC codes like
Osiris are vulnerable to numerical growth of emittance caused by the ``numerical Cherenkov effect'' \cite{godfrey:1974}.
This is a know issue with the Yee EMF solver which causes the phase velocity of electromagnetic fields to be lower than
$c$ while the beam moves very close to $c$. The effect can be mitigated somewhat by replacing the Yee solver with a
solver designed by Lehe \cite{lehe:2013}. Our simulations showed that this numerical effect is still prominent in the
high density regions of the electron beam, making it difficult to distinguish emittance growth from the physics from
that originating from numerical error.

Quasi-static PIC codes do not suffer this problem, so in order to study the emittance evolution of the beam we instead
turned to the recently released open source version QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D PIC
code \cite{huang:2006, an:2013}.

@@

In the QuickPIC simulations we use a single proton drive bunch that sets up a wakefield comparable to that which we
expect to see from the self-modulated SPS beam in AWAKE Run 2. The baseline AWAKE drive beam contains $3\nexp{11}$
protons \cite{gschwendtner:2016}. Only half of these are driving the wakefield, presuming the self-modulation is seeded
in the middle of the unmodulated bunch. In addition. a significant portion of the protons are lost during
self-modulation. The needed gap between the two plasma cells also contributes to a decreased density of protons near the
axis, and some of the field is drained by other protons seeing an accelerating phase within the drive beam itself. In
total, the expected peak $E_{z}$ field of $~2\funit{GV}{m}$ in the first plasma cell drops to $500-600\funit{MV}{m}$ in
the second cell \cite{awake_collaboration:2016}. A single proton bunch of $1.46\nexp{10}$, or $2.34\unit{nC}$ and
$7\unit{kA}$ was able to drive an $E_{z}$ field of a comparable magnitude in our single drive bunch setup.

The single bunch approach also has the benefit of reducing simulation time for the 3D simulations. Our aim was to create
a stable environment for the witness beam. We are not considering the evolution of the proton beam in this test setup,
so to prevent the proton beam from evolving radially, we increased the drive beam particle mass by a factor of
$1\mexp{6}$.

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. The
plasma electrons are depleted to around $65\%$ of nominal plasma density at the injection point of the electron beam in
reference simulations \cite{awake_collaboration:2016}. This condition is replicated in out single bunch case. The peak
density of the drive beam is $0.83\cdot n_{0}$ producing a quasi-linear condition behind it. The single bunch setup uses
the baseline proton energy $W_{pb} = 400\unit{GeV}$, and retains the transverse size $\sigma_{r} = 200\unit{\mu m}$
throughout the plasma cell. The length of the drive bunch $\sigma_{z} = 40\unit{\mu m}$.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. AWAKE baseline energy for the long witness beam for Run 1 is $W_{eb} = 10-20\unit{MeV}$
\cite{gschwendtner:2016}. In Run 2 the experiment will likely require a higher initial energy, but a
$W_{eb} \approx 50\unit{MeV}$ is sufficient to minimise de-phasing as the drift $\Delta\xi \propto \gamma^{-2}$
\cite{berglyd_olsen:2015}. For this case we have eliminated this source of de-phasing entirely in order to reduce the
number of variables.

Earlier simulations with witness beam radius $\sigma_{r} = 100\unit{\mu m}$ caused the beam to undergo rapid transverse
evolution as the beam entered the plasma region, reducing its $\sigma_{r}$ from $100\unit{\mu m}$ to $< 10\unit{\mu m}$
\cite{berglyd_olsen:2016}. This was primarily due to a mismatch between beam emittance and the plasma density. This
relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength.

In these simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{r}$ of $5.25\unit{\mu m}$. The relatively small size of the witness beam compared to the proton beam put some
restrictions on the transverse resolution of the witness beam. The transverse grid cell resolution was set to
$1.17\unit{\mu m}$.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive bunch is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. There is
some depletion of plasma electrons, but not enough to form a bubble. The dashed green line in the lower part of figure
\ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of $67\%$, close to what we see in full
scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

The witness bunch generates its own wakefield, which in the accelerating phase partially cancels out the $E_{z}$ field
generated by the drive bunch. For a finite length bunch this causes the particles in the tail of the witness bunch to be
accelerated less than those at the front resulting in higher energy spread in exchange for higher energy transfer
\cite{van_der_meer:1985}. Since $E_{z} \propto \rho_{b}$, a beam profile $\rho_{b}(\xi)$ that perfectly flattens the
accelerating field within the witness beam will reduces energy spread to zero. The ideal shape is triangular or
trapezoidal with the peak density in the direction of the beam velocity \cite{katsouleas:1987, tzoufras:2009}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    %~ \includegraphics[height=22.2mm,trim={ 0mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation01}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation02}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  0mm 0mm},clip]{figures/beamFilamentation03}
    %~ \includegraphics[height=22.2mm,trim={10mm 9.5mm 15mm 0mm},clip]{figures/beamOffsetMean}
    %~ \\
    %~ \includegraphics[height=26.0mm,trim={ 0mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation04}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation05}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  0mm 2mm},clip]{figures/beamFilamentation06}
    %~ \includegraphics[height=26.0mm,trim={10mm 0.1mm 15mm 2mm},clip]{figures/beamOffsetRMS}
    \caption{\label{Fig:BeamFilament} The left plots show the transverse phase space of the electron beam at different
        plasma positions. The Right side shows the macro particle mean position (top) and RMS spread (bottom). The blue
        particles and lines represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. The red
        particles and lines represent particles with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

For Gaussian beams a reasonable flat field, and consequently low energy spread, can be achieved by controlling the
charge density to achieve a similar effect. This, however, puts the front of the witness bunch outside of the ideal
region while the middle bulk of the beam has a trapezoidal-like shape. In our case, with a matched beam to the plasma
density defined by equation \ref{EQ:Matched}, we are limited on how much total charge we can accelerate without reaching
charge densities that will overload the field. The test case shown in figures \ref{Fig:BeamLoading} and
\ref{Fig:PlasmaDenTWake} has a total charge of $100\unit{pC}$ with a $\sigma_{r}=5.23\unit{\mu m}$ matching an initial
normalised emittance $\epsilon_{0} = 2\unit{\mu m}$.

As illustrated in figure \ref{Fig:BeamPS} our base case of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam, as
expected, shows a difference in energy transfer to the tail of the beam compared to the centre where the $E_{z}$-field
is nearly flat. Also the front of the beam has a large tail in momentum space as this regions sees a consistently
smaller accelerating field. The positioning of the witness beam is chosen to put the bulk of the charge as close to the
peak accelerating field as possible.

Due to the high charge density resulting from a very narrow beam, the witness beam's own wakefield reaches the full
non-linear regime with $n_{eb} \approx 35\times n_{0}$. This second wakefield is driven by the beam's own front. The
space charge rapidly becomes high enough to start expel plasma electrons from the beam axis, and form the characteristic
electrons sheet that defines the blowout regime. As a result the plasma region reaches full depletion close to the peak
of the electron beam, which in turn generates strong focusing fields, with gradients of $20\unit{kT/m}$ near the beam
axis (see figure \ref{Fig:PlasmaDenTWake}).

@@

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved. This has the potential to cause overloading of the field. However,
our $100\unit{pC}$ case is slightly under loaded, and this does not seem to be a significant effect in this instance.

@@

Our configuration also shows some robustness to small electron bunch transverse offsets as the bubble generated by the
witness bunch wakefield follows the bunch itself. The head of the bunch does not benefit from this effect, and we see
some defocusing in this region. This effect is likely to be greater for larger offsets as the bunch oscillates around
the axis of the drive bunch wakefield (see right hand plots of figure \ref{Fig:BeamFilament}). However, since the proton
wakefield creates a quasi linear condition, we still see the head of the beam stabilising after some time. The off-axis
case has a larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the growth levels
off after the first few metres and stays constant for the remainder of the plasma accelerator section (figure
\ref{Fig:BeamEmitt}).

@@

For accelerator applications in general, and also for Run 2 of the AWAKE experiment, it is desirable to maximise the
charge that can be successfully accelerated in the witness bunch. However, a longitudinally constant accelerating
field depends on both the position and the charge density of the witness beam \cite{katsouleas:1987, tzoufras:2009}.

@@

A higher charge beam will generate a non-linear wakefield closer to the head of the bunch, which in turns ensures that a
smaller portion of the beam experiences the quasi-linear region with emittance growth. As figure \ref{Fig:BeamQAbs}
illustrates, a higher charge also results in a lower energy gain.
